---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here
library(readxl)
library(janitor)
library(broom)
library(parsnip)
library(rsample)
library(tune)
library(recipes)
library(workflows)
library(ranger)
library(yardstick)
library(ggridges)

```


```{r load-data}
# load your data 
 

#Cleaning data 
BirdLife <- read.csv("data/AVONET1_BirdLife.csv") %>%
  filter(Order1 == "Passeriformes") %>%
  filter(!is.na(Habitat.Density))

BirdLife <- BirdLife %>%
  mutate(
    Habitat.Density = case_when(
      Habitat.Density == "1" ~ "Dense",
      Habitat.Density == "2" ~ "Open",
      Habitat.Density == "3" ~ "Open"
    ),
    Habitat.Density = fct_relevel(Habitat.Density, "Dense", "Open")
  )

#Summary statistics
BirdLife %>%
  group_by(Habitat.Density) %>%
  summarise(mean.Mass.Habitat = mean(Mass), mean.wing.length.habitat = mean(Wing.Length), sum.of.species = n_distinct(Species1))


bird_long <- BirdLife %>%
  pivot_longer(
    cols = c(Mass, Wing.Length),
    names_to = "Trait",
    values_to = "Value"
  ) %>%
  mutate(
    Value_plot = if_else(Trait == "Mass", log10(Value), Value)
  )

# Histogram
ggplot(BirdLife, aes(x = Mass, fill = Habitat.Density)) +
  geom_histogram(bins = 10) +
  scale_x_log10() +
 facet_wrap(~ Habitat.Density) + 
  labs(title = "Mass Distribution by Habitat Density (log scale)",
       x = "Mass (log10 g)",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

bird_long <- BirdLife %>%
  pivot_longer(
    cols = c(Mass, Wing.Length),
    names_to = "Trait",
    values_to = "Value"
  ) %>%
  mutate(
    Value_plot = if_else(Trait == "Mass", log10(Value), Value)
  )

#box plot
ggplot(bird_long, aes(x = Trait, y = Value, color = Habitat.Density)) +
  geom_boxplot(fill = NA, width = 0.6, outlier.shape = 1) +  
  facet_wrap(~ Habitat.Density) + 
  scale_y_log10() +   
  labs(title = "Mass and Wing.Length by Habitat Density",
       x = "Trait",
       y = "Value (log10 scale)") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none",
  )






BirdLife %>%
  ggplot(aes(x = (Wing.Length), 
                 y = (Mass), 
                 colour = Habitat, 
                 size = Total.individuals
             )) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  guides(size = FALSE, alpha = FALSE) +
  facet_wrap(. ~ Habitat.Density, scales = "free")

  
  
(result <- BirdLife %>%
  group_by(Habitat.Density) %>%
  group_modify( ~ {
    m <- linear_reg() %>%
      set_engine("lm") %>%
      fit(Wing.Length ~ Mass, data = .x)
  tidy(m)
  }))



# Each coefficient describes the expected difference between wing length in
# that particular habitat compeared to the baseline
linear_reg()%>%
  set_engine("lm") %>%
  fit(Mass ~ Habitat.Density, data = BirdLife) %>%
  tidy()
  
#Forest <- BirdLife %>%
  #filter(Habitat == "Forest")

wl_mass_fit <- linear_reg()%>%
  set_engine("lm") %>%
  fit(Wing.Length ~ log(Mass), data = BirdLife)

wl_mass_fit_aug <- augment(wl_mass_fit$fit)

ggplot(wl_mass_fit_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted wing length", y = "Residuals")


ggplot(BirdLife) +
  geom_histogram(aes(x = Mass)) +
  scale_x_log10() +
  facet_wrap(. ~ Habitat.Density, scales = "free")





#MODEL BUILDING
set.seed(1122)
BirdLife_split <- initial_split(BirdLife)

train_data <- training(BirdLife_split)
test_data <- testing(BirdLife_split)

  
BirdLife_rec <- recipe(
  Habitat.Density ~ Mass + Wing.Length + Beak.Width, 
  data = train_data
  ) %>%
  step_zv(all_predictors())

BirdLife_mod <- logistic_reg() %>%
  set_engine("glm")

BirdLife_wflow <- workflow() %>%
  add_model(BirdLife_mod) %>%
  add_recipe(BirdLife_rec)

BirdLife_fit <- BirdLife_wflow %>%
  fit(data = train_data)


(BirLife_train_pred <- predict(BirdLife_fit, train_data, type = "prob") %>%
  bind_cols(train_data))

BirLife_train_pred %>%
  roc_curve(
    truth = Habitat.Density,
    .pred_Dense,
    event_level = "first"
  ) %>%
  autoplot()

(summary(BirdLife_rec))


# MODEELLING WITH ALL PREDICTATORS



#step_dummy(all_nomial(), -all_outcomes()) %>%
# FOLDING?

#set.seed(345)
#folds <- vfold_cv(BirdLife, v = 5)
#set.seed(456)
#BirdLife_fit_rs <- BirdLife_wflow %>%
  #fit_resamples(folds)


```


