---
title: "The effect of habitat density on wing length and mass of different species of birds"
author: "by Tidybear123: Ishildur, Laith, Jason, & Richard"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(readxl)
library(janitor)
library(tibble)
library(gghighlight)
library(devtools)
library(readxl)
library(janitor)
library(broom)
library(parsnip)
library(rsample)
library(tune)
library(recipes)
library(workflows)
library(ranger)
library(yardstick)
library(ggridges)

```

```{r load-data, include=FALSE}
#Loading data

BirdLife_file <- read.csv("data/AVONET1_BirdLife.csv")

# Cleaning

BirdLife <- BirdLife_file %>%
  filter(Order1 == "Passeriformes", 
         !is.na(Habitat.Density),
         !is.na(Habitat))%>%
  mutate(
    Habitat.Density = case_when(
      Habitat.Density == 1 ~ "Dense",
      Habitat.Density == 2 | 3 ~ "Open",
    ),
    Habitat.Density = fct_relevel(Habitat.Density, "Dense", "Open")
  )


```

## Introduction

Bird habitats influence how birds move, forage, and fly, which can lead to differences in their physical traits. We wanted to explore whether habitat type affects key characteristics, with a particular focus on body mass. We found a paper that suggest that there is a link between body mass and habitat structure, especially when comparing dense and open environments. This motivated us to test whether the AVONET data set shows similar patterns. Our study focuses on Passeriformes, the largest and most diverse bird order, representing around 60% of all bird species. Based on this, we asked: 
What is the relationship between body mass and open vs. dense habitats in Passeriformes, and do other traits help explain where a species lives?

## Data Overview

For this project, we will be using the data set AVONET: morphological, ecological and geographical data for all birds. Each observation is a species of bird, containing information about its functional traits including ecological variables, continuous morphological traits, and information on range size and location. We chose to focus on Habitat, Habitat density, Mass and some other body measurements.

During our cleaning process we found that to get the sample sizes of dense and open habitats approximately equal we had to combine semi-open (2) and open (3) habitats to one value called "Open".

```{r}

table <- tribble(
  ~Variable,   ~Type,  ~Description,
  "Habitat.Density", "Discrete", "A number representing the density of the habitat with 1 being more dense and 3 being more open",
  "Mass", "Continous", "The average mass of the recorded birds in grams",
  "Habitat", "Categorical", "The average wing length of the recorded birds in centimeters"
)
knitr::kable(table)
```


Citation: 
AVONET. (2022). ‘AVONET: morphological, ecological and geographical data for all birds’. Available at: <https://figshare.com/s/b990722d72a26b5bfead>. Accessed: 9 November 2025

## Exploratory Statistics

We started by calculating the mean, median and standard deviation within each group. There we found that both the median and mean were smaller in Open Habitat. The 

```{r, echo = FALSE}
#Summary statistics
Summary_table <- BirdLife %>%
  group_by(Habitat.Density) %>%
  summarise(Mean = mean(Mass), Median = median(Mass), SD = sd(Mass), Total = n_distinct(Species1))

kable(Summary_table, digits = 3, format = "html")

```

## Body Mass Distrubution



```{r warning=FALSE, out.width="75%", fig.width=4, echo=FALSE}
ggplot(BirdLife, aes(y = Mass, x = Habitat.Density, color = Habitat.Density)) +
  geom_boxplot(fill = NA, width = 0.6, outlier.shape = 1) +  
  scale_y_log10() +   
  labs(title = "Mass by Habitat Density",
       x = "Type of Habitat",
       y = "Value (log10 scale)") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none",
  )
```



```{r}
BirdLife %>%
  ggplot(aes(y = Habitat), fill = Habitat) +
  geom_bar() +
  facet_wrap(~Habitat.Density) +
  labs(
    title = "Amount of Habitat Type in each Species",
    y = "Type of Habitat",
    x = "Number of species"
  )
```



### Predictive Modelling

```{r model building, echo = FALSE}

set.seed(222)
BirdLife_split <- initial_split(BirdLife)

train_data <- training(BirdLife_split)
test_data <- testing(BirdLife_split)

folds <- vfold_cv(train_data, v = 10)
  
rec_mass <- recipe(
  Habitat.Density ~ Mass,
  data = train_data
) %>%
  step_zv(all_predictors())

mod_glm <- logistic_reg() %>%
  set_engine("glm")

wflow_mass <- workflow() %>%
  add_model(mod_glm) %>%
  add_recipe(rec_mass)

cv_mass <- fit_resamples(
  wflow_mass,
  resamples = folds,
  metrics = metric_set(roc_auc),
  control = control_resamples(save_pred = TRUE)
)


rec_multi <- recipe(
  Habitat.Density ~ Mass + Wing.Length + Beak.Length_Culmen + Beak.Length_Nares
+ Beak.Width + Beak.Depth + Tarsus.Length + Wing.Length + Kipps.Distance + Tail.Length,
  data = train_data
) %>%
  step_zv(all_predictors())

wflow_multi <- workflow() %>%
  add_model(mod_glm) %>%
  add_recipe(rec_multi)

fit_multi <- wflow_multi %>%
  fit(data = train_data)

cv_multi <- fit_resamples(
  wflow_multi,
  resamples = folds,
  metrics = metric_set(roc_auc),
  control = control_resamples(save_pred = TRUE))


comperasion <- bind_rows(
  collect_metrics(cv_mass)  %>% mutate(model = "Mass only"),
  collect_metrics(cv_multi) %>% mutate(model = "All measurements")
) 

comperison_table <- comperasion %>%
  select(.metric, mean, std_err, model) %>%
  rename(
    Metric = .metric,
    Mean = mean,
    Std_Error = std_err,
    Model = model
  )

kable(comperison_table, digits = 3, format = "html")


```

## Limitations



## Summary



# the instructions


The write-up of your project and findings go here.

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation.
This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work.
Feel free to split this section into subsections to make it easier to read.

The length should be roughly 1,500 words.
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio.
Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good!
The addin will ignore code chunks and only count the words in prose.
If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.


## References

Tobias, J. A., Sheard, C., Pigot, A. L., M. Devenish, A. J., Yang, J., Sayol, F., C. Neate-Clegg, M. H., Alioravainen, N., Weeks, T. L., Barber, R. A., Walkden, P. A., A. MacGregor, H. E., I. Jones, S. E., Vincent, C., Phillips, A. G., Marples, N. M., Montaño-Centellas, F. A., Leandro-Silva, V., Claramunt, S., . . . Schleuning, M. (2022). AVONET: Morphological, ecological and geographical data for all birds. _Ecology Letters_, _25_(3), 581-597. https://doi.org/10.1111/ele.13898

Polo, V., & Carrascal, L. M. (1999). Shaping the body mass distribution of Passeriformes: Habitat use and body mass are evolutionarily and ecologically related. Journal of Animal Ecology, 68(2), 324-337. https://besjournals.onlinelibrary.wiley.com/doi/10.1046/j.1365-2656.1999.00282.x 

Open AI
