---
title: "The effect of habitat density on wing length and mass of different species of birds"
author: "by Tidybear123: Ishildur, Laith, Jason, & Richard"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-lib, include = FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(tibble)
library(gghighlight)

```

```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.


```

## Research Question

Describe your research question here.
Include why this question might be of interest.

the research we mainly want to focus on are the effect that the mass has on the wing length of a specie of bird and how this affect the density of the habitat that the birds live in, usually it's harder to stay afloat the heavier the object is, so birds tend to evolve to be as light as they can be, but for some predatory birds,

core idea: the heavier they are, the stronger their wings would have to be to support them to float in air

## Data Overview

for this project, we will be using the data set AVONET: morphological, ecological and geographical data for all birds, this database includes data about 90000 birds from across 11009 species of birds from all across the world, the data set includes data such as beak length, tail length, habitat and many more, for our project, we will be mainly looking at three variables, namely habitat density, mass and wing length

```{r}

table <- tribble(
  ~Variable,   ~Type,  ~Description,
  "Habitat.Density", "Numeric", "A number representing the density of the habitat with 1 being more dense and 3 being more open",
  "Mass", "Numeric", "The average mass of the recorded birds in grams",
  "Wing.Length", "Numeric", "The average wing lengrh of the recorded birds in centimeters"
)
knitr::kable(table)
```

we also created a variable called habitat density type to aid us in the research, because we felt that the details for habitat density were unnecessary for our project while we just want to see the effect of dense vs open habitats

```{r}

table2 <- tribble(
  ~Variable,   ~Type,  ~Description,
  "Habitat.Density.Type", "Character", "the habitat is considered open if the habitat density is 2 or 3, and considered dense if the habitat density is 1",
)
knitr::kable(table2)
```

Citation: AVONET. (2022). ‘AVONET: morphological, ecological and geographical data for all birds’. Available at: <https://figshare.com/s/b990722d72a26b5bfead>. Accessed: 9 November 2025

## Findings

### Wing Length against Mass

we first plotted a graph of the mass of birds against their wing length in an attempt to find any sort of pattern or trend, after we did this we realized something strange, some of the outliers in the data seems too extreme, while the majority of birds have mass below 11000, there are ten species with mass ranging from 23000 to 111000

```{r}
BirdLife <- read.csv("data/AVONET1_BirdLife.csv")
BirdLife %>%
  select(Mass,Wing.Length,Species1) %>%
  ggplot(aes(y = Wing.Length, x = Mass)) +
  geom_jitter() +
  labs(
    title = "bird mass against wing length",
    y = "wing length in cm",
    x = "mass in gram"
  )
```

this made the graph slightly harder to see and made us wonder as to why that might be the case, after some searching we found out that the ten outliers are mainly ostrich, cassowary, emus, penguins and rhea, all of them flightless,

```{r}
BirdLife %>%
  select(Mass,Wing.Length,Species1) %>%
  ggplot(aes(y = Wing.Length, x = Mass)) +
  geom_jitter() +
  gghighlight(Mass > 15000, label_key = Species1) +
  labs(
    title = "bird mass against wing length",
    y = "wing length in cm",
    x = "mass in gram"
  )
```

### Wing Length against Mass for birds under 15000 grams

since we are mainly looking at the effect that wing length have at the mass of the birds we decided that they won't matter as much since their weight are supported by their leg and being on ground rather than having stronger wings like the other birds, so we removed these outliers from our graph

```{r}
BirdLife %>%
  filter(Mass < 15000) %>%
  select(Mass,Wing.Length) %>%
  ggplot(aes(y = Wing.Length, x = Mass)) +
  geom_point() +
  labs(
    title = "bird mass against wing length",
    y = "wing length in cm",
    x = "mass in gram"
  )
```


# the instructions


The write-up of your project and findings go here.

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation.
This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work.
Feel free to split this section into subsections to make it easier to read.

The length should be roughly 1,500 words.
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio.
Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good!
The addin will ignore code chunks and only count the words in prose.
If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.


## References

List any references here.
You should, at a minimum, list your data source.
